Supervised Dimensionality Reduction with Random Projection-Based Ensemble Learning

### 2021-05-15 paper/main.tex

\subsection{False discovery and power for dimension reduction}
We use the definitions from \citet{taeb_false_2020} of false discovery and power for subspace discovery. In particular, given a target subspace $S^* \subset \mathbb{R}^p$, and an estimate subspace $\hat{S}$, we let the false discovery $FD$ be given by 
\[FD = \mathrm{Trace} \big( (I_p - \Pi_{S^*})\Pi_{\hat{S}} \big) \]
and power
\[PW = \mathrm{Trace} \big( \Pi_{S^*} \Pi_{\hat{S}} \big). \]
Here, $\Pi_{S^*}, \Pi_{\hat{S}}$ refer to projections onto $S^*$ and $\hat{S}$ respectively. 
\subsection{Random-projection ensemble classification}
We now describe random-projection ensemble classification as proposed by \citet{cannings_random-projection_2017}, a flexible tool which allows the use of a practitioner's favourite low-dimensional classifier. Random-projection ensemble classification will form the basis of our Let $X$ be a random vector taking values in $\mathcal{X} \subset \mathbb{R}^p$ from which we are attempting to predict a response $Y$ taking values in $\{0,1\}$. Furthermore suppose we are given $n$ observations of the pair, and denote our data $\mathcal{T}_n = \{(x_1, y_1), (x_2, y_2), \dots, (x_n, y_n)\}$. We choose a $d$-dimensional base classifier $C$, and for any projection $A:\mathbb{R}^p \rightarrow \mathbb{R}^d$ denote by $C^A$ that classifier applied to the projected observations $AX$ (so the prediction $C^A(x) = C(Ax)$). We write $\mathcal{T}_n^A$ for the projected dataset $\{(Ax_1, y_1), (Ax_2, y_2), \dots, (Ax_n, y_n)\}$. Then we may let $C^A_{\mathcal{T}_n^A}$ denote the classifier trained on the projected dataset and acting as a classifier on $\mathbb{R}^p$.

@@

\begin{algorithm}[H]
\SetAlgoLined
\KwResult{Output an ordered set of basis vectors with associated weightings}
 Choose parameters $d,B_1,B_2$\;
 \For{$b_1 = 1,2,\dots,B_1$}{
    \For {$b_2 = 1,2,\dots,B_2$}{
         Generate a random projection $ A_{b_1,b_2}:\mathbb{R}^p \rightarrow \mathbb{R}^d$\;
         Train the classifier $C^{A_{b_1,b_2}}_{\mathcal{T}_n}$\;
         Compute $R^{A_{b_1,b_2}}_{\mathcal{T}_n}$\;
        }
    Set $b_2^* := \sargmin\limits_{b_2 = 1,\dots, B_2} (R^{A_{b_1,b_2}}_{\mathcal{T}_n})$\;
    Set $A_{b_1} := A_{b_1,b_2^*}$\;
    }
    Compute $\Pi = \frac{1}{B_1}\sum\limits_{b_1}^{B_1}A_{b_1}^TA_{b_1}$\;
    Compute eigenvalue decomposition $\Pi = V\Lambda V^T$\;
    Output columns of $V$ as basis vectors\;
    Output diagonal elements of $\Lambda$ as weightings\;

@@

\section{Theoretical Analysis}
\tcnote{I think we should be able to say something about $\mathbf{E}(\Pi)$, where $\mathbf{E}$ denotes the average over the randomness in the projections only. And bound $\|\Pi - \mathbf{E}(\Pi)\|_{\mathrm{op}}$ and/or $\|\Pi - \mathbf{E}(\Pi)\|_{F}$ with high-probability. }

@@

% \subsection{Naive LDA }
% \tcnote{write out Linear Discriminant Analysis (and why naive?)}
% Suppose an LDA separation with independent covariates:
% $$X\ | \ Y = 0 \sim N_p(\mathbf{\mu}_0, \sigma^2 I_p) $$
% $$X\ | \ Y= 1 \sim N_p(\mathbf{\mu}_1,  \sigma^2 I_p) $$

@@

\section{Simulations}

@@

We present two simulation studies, of distributions with known central subspaces, and compare our method with other techniques for supervised dimension reduction. \\
~\\
\textbf{Sparse class boundaries}\\
First, we let 
\[X \ | \ \{Y = 0\} \sim \frac{1}{2}N_p(\mu_0, \Sigma) + \frac{1}{2}N_p(-\mu_0, \Sigma) \]
\[X \ | \ \{ Y = 1\} \sim \frac{1}{2}N_p(\mu_1, \Sigma) + \frac{1}{2}N_p(-\mu_1, \Sigma). \]
We set $p=100$, $\mu_0 = (2, -2, 0, \ldots, 0)^T$, $\mu_1 = (2, 2, 0, \ldots, 0)^T$ and let $\Sigma = I_{100}$. This model exhibits a setup in which a central subspace exists (of dimension $d^* = 2)$, but there is no good linear class boundary. \\
~\\

\begin{table}
  \caption{Simulation performance}
  \label{sample-table}
  \centering
  \begin{tabular}{l|lll|lll}
    \toprule
    \multicolumn{2}{c}{Model 1} & \multicolumn{2}{c}{Model 2}                     \\
    \cmidrule(r){1-4}
    Method     & Power     & False & Accuracy &  Power     & False & Accuracy &  \\
    & & Discovery & & & Discovery &
    \midrule
    RP-LDA & Input terminal  & $\sim$100  &  &&& \\
    RP-QDA    & Output terminal & $\sim$10    &  &&&\\
    RP-kNN     & Cell body       & up to $10^6$ & &&& \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\textbf{Rotated sparse normal}\\
Here,

>>>>

\subsection{False discovery and power for dimension reduction}
We use the definitions from \citet{taeb_false_2020} of false discovery and power for subspace discovery. In particular, given a target subspace $S^* \subset \mathbb{R}^p$ with $dim(S^*) = d^*$, and an estimate subspace $\hat{S}$, we let the false discovery rate $FDR$ be given by 
\[FDR = \frac{\mathrm{Trace} \big( (I_p - \Pi_{S^*})\Pi_{\hat{S}} \big)}{p - d^*} \]
and power
\[PW = \frac{\mathrm{Trace} \big( \Pi_{S^*} \Pi_{\hat{S}} \big)}{d^*}. \]
Here, $\Pi_{S^*}, \Pi_{\hat{S}}$ refer to projections onto $S^*$ and $\hat{S}$ respectively. 
\subsection{Random-projection ensemble classification}
We now describe random-projection ensemble classification as proposed by \citet{cannings_random-projection_2017}, a flexible tool which allows the use of a practitioner's favourite low-dimensional classifier. Random-projection ensemble classification will form the basis of our Let $X$ be a random vector taking values in $\mathcal{X} \subset \mathbb{R}^p$ from which we are attempting to predict a response $Y$ taking values in $\{0,1\}$. Furthermore suppose we are given $n$ observations of the pair, and denote our data $\mathcal{T}_n = \{(x_1, y_1), (x_2, y_2), \dots, (x_n, y_n)\}$. We choose a $d$-dimensional base classifier $C$, and for any projection $A:\mathbb{R}^p \rightarrow \mathbb{R}^d$ denote by $C^A$ that classifier applied to the projected observations $AX$ (so the prediction $C^A(x) = C(Ax)$). We write $\mathcal{T}_n^A$ for the projected dataset $\{(Ax_1, y_1), (Ax_2, y_2), \dots, (Ax_n, y_n)\}$. Then we may let $C^A_{\mathcal{T}_n^A}$ denote the classifier trained on the projected dataset and acting as a classifier on $\mathbb{R}^p$.

@@

\begin{algorithm}[H]
\SetAlgoLined
\KwResult{Output an ordered set of basis vectors with associated weightings}
 Choose parameters $d,B_1,B_2$\;
 \For{$b_1 = 1,2,\dots,B_1$}{
    \For {$b_2 = 1,2,\dots,B_2$}{
         Generate a random projection $ A_{b_1,b_2}:\mathbb{R}^p \rightarrow \mathbb{R}^d$\;
         Train the classifier $C^{A_{b_1,b_2}}_{\mathcal{T}_n}$\;
         Compute $R^{A_{b_1,b_2}}_{\mathcal{T}_n}$\;
        }
    Set $b_2^* := \sargmin\limits_{b_2 = 1,\dots, B_2} (R^{A_{b_1,b_2}}_{\mathcal{T}_n})$\;
    Set $A_{b_1} := A_{b_1,b_2^*}$\;
    }
    Compute $\Pi = \frac{1}{B_1}\sum\limits_{b_1 = 1}^{B_1}A_{b_1}^TA_{b_1}$\;
    Compute eigenvalue decomposition $\Pi = V\Lambda V^T$\;
    Output columns of $V$ as basis vectors\;
    Output diagonal elements of $\Lambda$ as weightings\;

@@

\section{Theoretical Analysis}

\tcnote{I think we should be able to say something about $\mathbf{E}(\Pi)$, where $\mathbf{E}$ denotes the average over the randomness in the projections only. And bound $\|\Pi - \mathbf{E}(\Pi)\|_{\mathrm{op}}$ and/or $\|\Pi - \mathbf{E}(\Pi)\|_{F}$ with high-probability. }

@@

\subsection{Linear Discriminant Analysis}
Suppose we have the following (canonical) model:
\[
X | \{Y = 1\} \sim N_p(\Delta/2 e_1, I_p); \quad X | \{Y = 0\} \sim N_p(-\Delta/2 e_1, I_p),
\]
where $e_1 = (1, 0, \ldots, 0)$ is the first Euclidean basis vector and $I_p$ denotes the $p$ by $p$ identity matrix, and $\pi_1 = \mathbb{P}(Y=1) = \mathbb{P}(Y=0) = 1/2$. The Bayes risk in this case is $\Phi(-\Delta/2)$, where $\Phi$ denotes the standard Normal distribution function.

Fix a ``projection" matrix $A: \mathbb{R}^p \rightarrow \mathbb{R}^d$, and write $A = (a_1, \ldots, a_p)$ for the columns of $A$. We have 
\[
AX | \{Y = 1\} \sim N_p(\Delta/2 a_1, AA^T); \quad AX | \{Y = 0\} \sim N_p(-\Delta/2 a_1, AA^T). 
\] 
Then the \emph{projected} Bayes risk is 
\[
R(C_{A}^{\mathrm{Bayes}}) = \Phi\Bigl(-\frac{\Delta}{2} \|(AA^T)^{-1/2} a_1 \| \Bigr).
\]
Heuristically, we see that good projections have $\|(AA^T)^{-1/2} a_1\|$ close to 1.  Suppose now that $d =1$, then $\|(AA^T)^{-1/2} a_1\| = \frac{|a_1|}{\|A\|}$. Note also that 
\[   
PW(A)   = \mathrm{tr}(\Pi_{e_1}\Pi_{A}) = \mathrm{tr}\Bigl(e_1e_1^T\frac{A^TA}{\|A\|^2}\Bigr)  =  \frac{a_1^2}{\|A\|^2}.
\]

Now, for $B_2 \geq 1$, let $\mathbf{A}_1^T, \ldots, \mathbf{A}_{B_2}^T \stackrel{\mathrm{i.i.d}}{\sim} N_p(0, \frac{1}{p} I_p)$. Now write
\[
 \frac{|a_1|}{\|A\|} = \frac{A^T e_1 } {\|A\|} = \cos(\Theta(A^T, e_1)). 
\] 
We also have that $\frac{A^T}{\|A\|}$ is the leading eigenvector of $A^TA$, interpreting $e_1$ as the leading eigenvector of $e_1e_1^T$, we have that 
\begin{align*}
\sin\bigl(\Theta(A^T, e_1)\bigr) &\leq 2\Bigl\|\frac{A^TA}{\|A\|^2} -  e_1e_1^T\Bigr\|_{\mathrm{op} } 
\\ & = \frac{ 2}{\|A\|^2} \bigl\| A^TA - \|A\|^2 e_1e_1^T \bigr\|_{\mathrm{op}} 
\\ & = \frac{2}{\|A\|^2} \| (A^T - \|A\|e_1)(A + \|A\|e_1^T)\|_{\mathrm{op}} 
\\ & = 2\Bigl\|\frac{A^T}{\|A\|} - e_1\Bigr\|\Bigl\|\frac{A^T}{\|A\|} + e_1\Bigr\| \leq 4 \Bigl\|\frac{A^T}{\|A\|} - e_1\Bigr\|.  
\end{align*}

We are interested in 
\begin{enumerate} 
\item the distribution of 
\[
\min_{b_2 = 1, \ldots, B_2} \{ R(C_{\mathbf{A}_{b_2}}^{\mathrm{Bayes}})\} 
\]
\item the quantity
\[
\hat{R}_{\mathbf{A}_{b_2}} - R(C_{\mathbf{A}_{b_2}}^{\mathrm{Bayes}})
\]
where $\hat{R}$ is some estimator of the risk. 
\item the quantity
\[
\min_{b_2 = 1, \ldots, B_2} \hat{R}_{\mathbf{A}_{b_2}} - \min_{b_2 = 1, \ldots, B_2} R(C_{\mathbf{A}_{b_2}}^{\mathrm{Bayes}})
\]
\item the distribution of $\mathbf{A}_{b^*_2}$, where 
\[
b^*_2 = \argmin_{b_2 = 1, \ldots, B_2}  \{ R(C_{\mathbf{A}_{b_2}}^{\mathrm{Bayes}})\} 
\]
\item the distribution of $\mathbf{A}_{\hat{b}_2}$, where 
\[
\hat{b}_2 = \argmin_{b_2 = 1, \ldots, B_2} \hat{R}_{\mathbf{A}_{b_2}} 
\]
\end{enumerate}

Suppose we have access to $R(C_{A}^{\mathrm{Bayes}})$, for every $A$, then we will pick the projection with the largest value of $\frac{a_1^2}{\|A\|^2}$. In other words 
\[
\argmin_{b_2 = 1, \ldots, B_2} \{ R(C_{\mathbf{A}_{b_2}}^{\mathrm{Bayes}})\} = \argmax_{b_2 = 1, \ldots, B_2} \{\cos^2(\Theta(\mathbf{A}_{b_2}^T, e_1))\} 
\]

@@

\section{Simulations}
\subsection{Risk versus discovery: an illustrative example}
It is natural to ask under what circumstances we may recover a classifier that performs very well

\begin{figure}[htbp]
    \centering
    \includegraphics[width=4in]{../results/figures/ChangingDelta.png}
    \caption{\textbf{A}: Component-aligned, axis-aligned; \textbf{B}: Not component-aligned, axis-aligned; \textbf{C}: Component-aligned, not axis-aligned; \textbf{D}: Not component-aligned, not axis-aligned.\label{fig:examplefigures}}
\end{figure}

@@

We present two simulation studies, of distributions with known central subspaces, and compare our method with other techniques for supervised dimension reduction. \\
~\\
\textbf{Model 0: 'Doughnuts in Two Dimensions'} \\
We consider random noise in 8 components, and concentric rings of membership to classes 0 and 1 in the first two dimensions (see figure). This ensures that within-class means and covariances are all matched. \\

\begin{figure}[htbp]
    \centering
    \includegraphics[width=4in]{../results/figures/doughnuts.png}
    \caption{The first four components of data from model 0 (two signal, two noise).\label{fig:doughnuts}}
\end{figure}

\begin{table}
  \caption{Simulation performance}
  \label{sample-table}
  \centering
  \begin{tabular}{l|lll}
    \toprule
    & \multicolumn{3}{c}{Model 0}                  \\
    \cmidrule(r){1-4}
    Method     & Power     & FDR & Accuracy  \\

    \cmidrule(r){1-4}
    RP-LDA & 0.0510 & 0.237  &   0.490\\
    RP-QDA & 0.453 & 0.137 &  0.734 \\
    RP-kNN & 0.994  & 0.00147 & 0.999\\
    SIR & - &  - &  -  \\
    SAVE & 0.164 & 0.210 & 0.470 \\
    PHD & 0.172 & 0.207 &     0.502 \\
    PCA & 0.207 & 0.198 & 0.508 \\
    \bottomrule
  \end{tabular}
\end{table}

~\\

\noindent\textbf{Model 1: Sparse class boundaries}\\
We let 
\[X \ | \ \{Y = 0\} \sim \frac{1}{2}N_p(\mu_0, \Sigma) + \frac{1}{2}N_p(-\mu_0, \Sigma) \]
\[X \ | \ \{ Y = 1\} \sim \frac{1}{2}N_p(\mu_1, \Sigma) + \frac{1}{2}N_p(-\mu_1, \Sigma). \]
We set $p=100$, $\mu_0 = (2, -2, 0, \ldots, 0)^T$, $\mu_1 = (2, 2, 0, \ldots, 0)^T$ and let $\Sigma = I_{100}$. This model exhibits a setup in which a central subspace exists (of dimension $d^* = 2)$, but there is no good linear class boundary. We have Bayes risk given by $2\Phi(2)\Phi(-2)$, and projected risk given by 
\[ PBR(A) = \mathbb{P}\Big( \big((\Sigma^{\frac{1}{2}}Z) \cdot n_1 > - \mu_0 \cdot n_1 \big)  \cap \big( (\Sigma^{\frac{1}{2}}Z) \cdot n_2 > - \mu_0 \cdot n_2 \big) \Big) \]
where 
\begin{align*}
    n_1 & = A^T(A\Sigma A^T)^{-1}A(\mu_1 - \mu_0) \\
    n_2 & = A^T(A \Sigma A^T)^{-1} (\mu_0 + \mu_1) \\
    Z & \sim N(0, I_p).
\end{align*}

~\\

\begin{table}
  \caption{Simulation performance}
  \label{sample-table}
  \centering
  \begin{tabular}{l|lll|lll}
    \toprule
    & \multicolumn{3}{c}{Model 1} & \multicolumn{3}{c}{Model 2}                     \\
    \cmidrule(r){1-7}
    Method     & Power     & FDR & Accuracy &  Power     & FDR & Accuracy   \\

    \midrule
    RP-LDA & 0.0354 & 0.0197  &  0.495 & 0.321 &   0.0213 & 0.963 \\
    RP-QDA & 0.0639  & 0.0191 & 0.495 & 0.307 &    0.0214 & 0.950 \\
    RP-kNN & 0.00621  & 0.0203 & 0.470 & 0.324 &   0.0209 & 0.959 \\
    SIR & 0.493 &  0.0103 & 0.650 &  0.0903 &      0.0281 & 0.751 \\
    SAVE & 0.931 & 0.00140 & 0.878 & 0.0983 &      0.0279 & 0.717 \\
    PHD & 0.896 & 0.00211 & 0.869 & 0.0353 &       0.0298 & 0.654 \\
    PCA & 0.0132 & 0.0201 & 0.493 & 0.619 &        0.0118 & 0.950 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\noindent\textbf{Model 2: Rotated sparse normal}\\
Here,
